# Regression analysis of the IODS project 

*This section describes the steps taken to perform regression analysis of data obtained for this IODS project.*

First, let me begin by reading the data for this analysis into R from my local computer. *Please note that the data here was pre-processed throuhg Wrangling from the Helsinki Open data science project, which originally can be  found in this link http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS3-data.txt*   

```{r}
# install.packages("ggplot2")
library(ggplot2)
# Reading the data from my local computer
data <- read.csv("C:\\Users\\fridaya\\Documents\\IODS-project\\data\\learning2014.csv")
learning2014 <- data
```
Now that we have successfully loaded the data to R, we would like to explore the structure and dimension of the data

```{r}
# Here we are checking the structure of the data learning2014
str(learning2014)
```
```{r}
# Here we are checking the dimension of the data learning2014
dim(learning2014)
```
Now we would like to show the graphical overview of the data 

```{r}
# Here, we try to show the graphical entire overview of the data using plot

plot(learning2014)
```
Figure 1

In Figure 1  above, we see the overview of the variables (gender, age, attitude, deep, stra, surf, and points) in the learning2014 data frame. 

*Note* that the variable "gender" scattered plot is not consistent with the other variables. This could be becuase the gender variable did not contain numeric values.

```{r}
# Here, we try to show the graphical overview of the data using pairs excluding the first variable 
pairs(learning2014[-1])
```
Figure 2.

Notice that we repeated the ploting of the variables in learning2014 data by using the pairs function and excluding "gender". 


*interpretations:*

Figure 2 presents the selected variables in diagonal. each row displays a variable on the y-axis and the other variables on the corresponding x- axis. For example, the first row "age" is displayed on the y-axis whereas attitude, deep, stra, and surf are displayed on the x-axis for each of the 5 plots on that row. 


```{r}
# Here we are using three variables to fit our regression model. The variables are attitude, and points, where points is out dependent variable.

library(ggplot2)
qplot(attitude, points, data = learning2014) + geom_smooth(method = "lm")

```

```{r}

# Here we try to fit our a linear model. The dependent variable is points while the explanatory variable is attitude
model <- lm(points ~ attitude, data = learning2014)

# Now let us print the summary of our model
summary(model)
```


*Explanation of the model summary*



```{r}


```

