# Analysis of longitudinal data (Chapter 6) 


*We begin in this chapter by reading the BPRS and RATS  wide form and the converted BPRSLONG and RATSLONG long form already completed and saved during the data wrangling exercise in "meet_and_repeat" R script.*


```{r}
library(dplyr)
library(ggplot2)
library(GGally)
library(tidyr)
```


### Summary of BPRS Data 

```{r}
# Let us read the BPRS data and take a look at the data summary
BPRS  <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep=" ", header=TRUE) # BPRS Wide form

# Let us check the BPRS data names, structures, and summary 

names(BPRS) # ... checking the column names 
str(BPRS) # ... checking the structure of the BPRS 
summary(BPRS) # .... checking the summary of the BPRS data
#BPRS

```

The brief psychiatric rating scale (BPRS) is used to evaluate whether or not the patient has schizophrenia. The data contain 40 observations and 11 variables.


```{r}
# Next, we load the long form of the BPRS saved in the local computer 
BPRSLONG <- read.csv("C:\\Users\\fridaya\\Documents\\IODS-project\\data\\BPRSLONG.csv")

# Let us check the BPRSLONG data names, structures, and summary 

names(BPRSLONG) # ... checking the column names 
str(BPRSLONG) # ... checking the structure of the BPRSLONG data
summary(BPRSLONG) # .... checking the summary of the BPRSLONG data


```

We observed that the Long form of the data contained 360 observations and 5 variables. 


### Summary of RATS data 

```{r}
# Let us read the RATS data and take a look at the data summary
RATS  <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", sep="\t", header=TRUE) # RATS Wide form

# Let us check the RATS data names, structures, and summary 

names(RATS) # ... checking the column names 
str(RATS) # ... checking the structure of the RATS 
summary(RATS) # .... checking the summary of the RATS data
#BPRS

```

The RATS data and their nutrition in its wide form  contain 16 observations and 13 variables

```{r}
# Next, we load the long form of the RATS data saved in the local computer 
RATSLONG <- read.csv("C:\\Users\\fridaya\\Documents\\IODS-project\\data\\RATSLONG.csv")

# Let us check the RATSLONG data names, structures, and summary 

names(RATSLONG) # ... checking the column names 
str(RATSLONG) # ... checking the structure of the RATSLONG 
summary(RATSLONG) # .... checking the summary of the RATSLONG data



```

After converting the RATS data to its long form, we then have 176 observations and 5 variables.


### Analyses of Chapter 8 of MABS using the RATS data (7 points) 

```{r}
# We now want to make the ID and Group variables in RATSLONG both factors instead of integers
RATSLONG$ID <- factor(RATSLONG$ID)
RATSLONG$Group <- factor(RATSLONG$Group)

# Lets check the structure of RATSLONG now
str(RATSLONG)

```



```{r}
# Draw the individual plot
ggplot(RATSLONG, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSLONG$Weight), max(RATSLONG$Weight)))


``` 


As we can see from the graph, the groups clearly have different values of weight. The first group seems to have the lowest weight while the last group has the highest weight

Next, we will like to standardize the data RATSLONG and re-plot the graph to see if there would be difference.

```{r}
# Standardizing the the variable Weight in the RATSLONG data 
RATSLONG <- RATSLONG %>%
  group_by(Time) %>%
  mutate(stdWeight = Weight) %>%
  ungroup()
# Let us glimpse the RATSLONG data to see the effect of our standardization
glimpse(RATSLONG)


# Now, let us re-draw the same  plot

ggplot(RATSLONG, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized bprs")
``` 

The differences in these two plots could be hard to deduce, however, it can be seen that a new feature (ID) has been added to the plot


NOw we want to make a summary data 

```{r}
# Number of subjects (per group):
n <- 20
# Make a summary data:
RATSS <- RATSLONG %>%
  group_by(Group, Time) %>%
  summarise( mean=mean(Weight), se=sd(Weight)/sqrt(n) ) %>%
  ungroup()

# Let us take the glimpse of the RATSS 

glimpse(RATSS)

```       

Now, we would draw a boxplot the 

```{r}
# Plotting the mean profiles 
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.9,0.4)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")

```

The plot of mean profiles clearly shows that there are no overlaping between the groups. This implies that the groups are different


Next, we will Create a summary data by group 

```{r}
# creating a summary data by Group and ID with mean as the summary variable (excluding the baseline)
RATSL8S <- RATSLONG %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Lets glimpse the data
glimpse(RATSL8S)


# plotting a boxplot of the mean versus group
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight)")


```

The boxplot shows some outliers. Particuarly in the second group, there is a value that is very high  compared to the other groups value, which creats an outlier.


#### Let us try to remove this outlier in the second group and see the behaviour of the graph.

```{r}
# Let us use the filter method and reset the condition
# create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
RATSL8S1 <- RATSL8S %>% filter(mean < 550)

ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), 64 days")

```

As shown in the redrawn plot, the outlier was removed that the group align more properly.

#### Now we need to perform formal test to show that the groups are truely different
```{r}

# Add the baseline from the original data as a new variable to the summary data
RATSL8S2 <- RATSL8S %>%
  mutate(baseline = RATS$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATSL8S2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)

```

From the analysis of variance, the group has very slightly significant correlation (0.076, p<0.001***) 



### Analyses of Chapter 9 of MABS using the BPRS  Data  (8 points) 
```{r}
# Let us check the structure of the BPRS and BPRSLONG data again 

str(BPRS) 

str(BPRSLONG)

# making factors out of treatment and subject variables of the Long form 
BPRSLONG$treatment <- factor(BPRSLONG$treatment)
BPRSLONG$subject <- factor(BPRSLONG$subject)

str(BPRSLONG)
```

Now the "treatment" and "subject" variables are now factors

```{r}
# summaries of the variables
summary(BPRSLONG)

```

Let us now plot the BPRSLONG data

```{r}
# Plot the BPRSLONG data
ggplot(BPRSLONG, aes(x = week, y = bprs, group = treatment)) +
  geom_line()

#graph showing all the data
ggplot(BPRSLONG, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSLONG$bprs), max(BPRSLONG$bprs)))
```

Let us create a regression model for the data
```{r}
# create a regression model RATS_reg
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSLONG)

# print out a summary of the model
summary(BPRS_reg)

```




Let us create the random intercept model
```{r}
# access library lme4
library(lme4)

# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSLONG, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)
```

```{r}

# create a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSLONG, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)
```

#### Creating the random intersept and random slope model 

```{r}
# create a random intercept and random slope model with the interaction
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | treatment), data = BPRSLONG, REML = FALSE)

# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)

# draw the plot of BPRS with the observed Weight values
ggplot(BPRSLONG, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_x_continuous(name = "Weeks") +
  scale_y_continuous(name = "BPRS") +
  theme(legend.position = "none") +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both)

```

```{r}

# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref1)

# create a new column fitted to RATSL
BPRSLONG <- BPRSLONG %>% mutate(fitted = Fitted)

# draw the plot of BPRS with the observed Weight values
ggplot(BPRSLONG, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_x_continuous(name = "Weeks") +
  scale_y_continuous(name = "BPRS") +
  theme(legend.position = "none") +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both)

```



